A pandemic-scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-09-28 latex/main.tex
Pass through the section on high-level results and dels

Using sc2ts, we built an ARG from 2,482,157 samples from the Viridian v04
dataset\cite{Hunt2024}, collected between 2020-01-01 and 2023-02-20.
Inference took 21 days on a XX core server with 500 GB of RAM 
($\sim 5$ CPU years; Figure~\ref{fig:inference_resources}).
After postprocessing, the ARG 
contains 2,689,054 nodes,  % Note the estimated dates 
2,748,838 edges, and 2,285,344 mutations at 29,893 sites. 
Including sample IDs and the Pango and Scorpio designations computed by
Pangolin~\cite{OToole2021assignment} as metadata for each node,
the ARG is stored in a 32MiB tszip format file.
% It's confusing to talk about recombinant QC here as we're talking 
% about the full ARG for the rest of the section.
[Do we need to talk about recombination?]
Of the nodes, 855 represent recombination events, 
resulting in 316 distinct trees along the genome (some
events occured at the same breakpoints [vague, rephrase]).
[TODO note about loading up the time it takes to get the node 
data dataframes in a notebook?]

% PARA: we compared against the UShER tree in a bunch of ways
% NOTE: Trying to clarify that this of the pandemic period and we're 
% not trying to cover *all* history.
The ARG is a comprehensive record of the pandemic phase 
of SARS-CoV-2 evolutionary history, and accurately reflects
a range of results from the literature.
First, we compared the sc2ts ARG and with the state-of-the-art UShER 
tree inferred from the same dataset
by pruning both to their intersection of 2,475,418 samples and 27,507 sites
(STAR Methods)
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
and found they were very similar in terms of parsimony
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We also compared the mutational spectra of major VOCs in the ARG
against the UShER tree and 
a previous study \cite{Bloom2023} (STAR Methods),
finding close agreement (Figure~\ref{fig:mutational_spectra}).
We validated that the broad phylogenetic structure of the ARG and 
UShER tree closely match by generating tanglegrams for a set of representative
samples from 1097 deeply-sampled lineages (STAR Methods; 
Figures~\ref{fig:tanglegram_base},
\ref{fig:tanglegram_delta},
\ref{fig:tanglegram_ba2},
\ref{fig:tanglegram_ba5}).

Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
but because they are difficult to disintinguish from bioinformatics errors 
and their multi-base nature presents challenges to the Markovian structure of 
the LS HMM, we masked out all gap characters as missing data during primary
inference. To incorporate the most important deletions into the ARG, we 
remapped data using posthoc parsimony at 75 sites involved in deletions 
with a frequency of $\geq 1\%$ reported by Li et al.\cite{Li2023indels}
during postprocessing (STAR Methods).
Although sites are treated independently in this approach,
runs of mutations to the gap character at adjacent sites
correctly identified many known 
high frequency (Table~\ref{tab:major_dels}) and 
recurrent (Table~\ref{tab:rec_dels}) deletions.
For example, the 4 characteristic deletions associated with 
% FIXME: this is just my notation - is this something recognised more 
% widely? If not, fix to something that is?
Alpha (11288+9, 21765+6, 21991+3, 28271+1) are correctly
identified as having a single origin [CITATION?].
The 21765+6 deletion (Spike 69-70)
which causes S-Gene Target Failure \cite{Walker2021},
emerges two more times at the origins of the 
BA.1 and BA.4 variants.
Although most highly recurrent deletions are likely errors,
some known truly recurrent deletions are present,
for example, two deletions in S1-NTD [WHICH TWO???]
associated with antibody escape \cite{McCarthy2021}
See Document\ref{sec:supplement_dels} for further analysis.

caused by sites enriched for errors [citation?],
sc2ts masks out a set of 100 sites identified
as the most highly homoplasic
by running on a subset of the data (STAR Methods).
Of these sites, 9 are in the set of [1XX] ``problematic sites''
masked by UShER \cite{Hunt2024}.
In addition to remapping the data for sites identified
as being involved in deletions,
we added mutations for XX of these highly recurrent sites
that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony.
Each node in the ARG was then assigned

to one of 2058 Pango lineages using pangolin (STAR Methods).
All but 21,680 sampled sequences (0.87\%) agreed
with the source sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data
versions 1.21 and 1.29 in the source Viridian metadata).
A majority of these lineages (1473 of 2058)
have monophyletic origins within the ARG,
increasing to 1779 (86\%) when we allow
for multiple sibling origination nodes
(Document~\ref{sec:supplement_pango_lineages}).

>>>>

Using sc2ts, we built an ARG from 2,482,157 samples from the Viridian v04
dataset\cite{Hunt2024}, collected between 2020-01-01 and 2023-02-20.
Inference took 21 days on a XX core server with 500 GB of RAM 
($\sim 5$ CPU years; Figure~\ref{fig:inference_resources}).
After postprocessing, the ARG 
contains 2,689,054 nodes,  % Note the estimated dates ?
2,748,838 edges, and 2,285,344 mutations at 29,893 sites. 
Including sample IDs and the Pango and Scorpio designations computed by
Pangolin~\cite{OToole2021assignment} as metadata for each node,
the ARG is stored in a 32MiB tszip format file.
% It's confusing to talk about recombinant QC here as we're talking 
% about the full ARG for the rest of the section.
Of the nodes, 855 represent recombination events, 
resulting in 316 distinct trees along the genome (some
events occured at the same breakpoints [vague, rephrase]).
[TODO note about loading up the time it takes to get the node 
data dataframes in a notebook?]

% PARA: we compared against the UShER tree in a bunch of ways
% NOTE: Trying to clarify that this of the pandemic period and we're 
% not trying to cover *all* history.
The ARG is a comprehensive record of the pandemic phase 
of SARS-CoV-2 evolutionary history, and accurately reflects
a range of results from the literature.
First, we compared the sc2ts ARG and with the state-of-the-art UShER 
tree inferred from the same dataset
by pruning both to their intersection of 2,475,418 samples and 27,507 sites
(STAR Methods)
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
and found they were very similar in terms of parsimony
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We also compared the mutational spectra of major VOCs in the ARG
against the UShER tree and 
a previous study \cite{Bloom2023} (STAR Methods),
finding close agreement (Figure~\ref{fig:mutational_spectra}).
We validated that the broad phylogenetic structure of the ARG and 
UShER tree closely match by generating tanglegrams for a set of representative
samples from 1097 deeply-sampled lineages (STAR Methods; 
Figures~\ref{fig:tanglegram_base},
\ref{fig:tanglegram_delta},
\ref{fig:tanglegram_ba2},
\ref{fig:tanglegram_ba5}).
We also evaluated how well the ARG reflects the structure 
of the Pango nomenclature system~\cite{Rambaut2020} by computing 
Pango assignments for all 2.7 million nodes in the ARG using 
Pangolin~\cite{OToole2021assignment} (STAR Methods).
All but 21,680 of the exported sample alignments (0.87\%) agreed
with the source consensus sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data
versions 1.21 and 1.29 in the source Viridian metadata).
A majority of these lineages (1473 of 2058)
have monophyletic origins within the ARG,
increasing to 1779 (86\%) when we allow
for multiple sibling origination nodes
(Document~\ref{sec:supplement_pango_lineages}).

[TODO we should mention UShER here - what do they do with dels?]
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
but because they are difficult to disintinguish from bioinformatics errors 
and their multi-base nature presents challenges to the Markovian structure of 
the LS HMM, we masked out all gap characters as missing data during primary
inference. 
To incorporate the most important deletions into the ARG, we 
remapped data using posthoc parsimony at 75 sites involved in deletions 
with a frequency of $\geq 1\%$ reported by Li et al.\cite{Li2023indels}
during postprocessing (STAR Methods).
Although sites are treated independently in this approach,
runs of mutations to the gap character at adjacent sites
correctly identified many known 
high frequency (Table~\ref{tab:major_dels}) and 
recurrent (Table~\ref{tab:rec_dels}) deletions.
For example, the 4 characteristic deletions associated with 
% FIXME: this is just my notation - is this something recognised more 
% widely? If not, fix to something that is?
Alpha (11288+9, 21765+6, 21991+3, 28271+1) are correctly
identified as having a single origin [CITATION?]
(Figure~\ref{fig:subgraph_alpha}).
The 21765+6 deletion (Spike 69-70)
which causes S-Gene Target Failure \cite{Walker2021},
emerges two more times at the origins of the 
BA.1 and BA.4 variants.
Although most highly recurrent deletions are likely errors,
some known truly recurrent deletions are present,
for example, two deletions in S1-NTD [WHICH TWO???]
associated with antibody escape\cite{McCarthy2021}.
See Document\ref{sec:supplement_dels} for further analysis.
